\documentclass[12pt,letterpaper]{report}

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{color}
\usepackage{enumitem}
\usepackage{epsfig}
\usepackage{framed}
\usepackage[letterpaper, margin=1in]{geometry}		% Margins should be 1 inch according to McGill requirements (https://www.mcgill.ca/gps/thesis/guidelines/preparation)
\usepackage{graphicx}
\usepackage{listings}
\usepackage{placeins}
\usepackage{setspace}
\usepackage{subcaption}

\input{def_mcgill}

%-----------------------------------------------------------------------------
% Student info
%-----------------------------------------------------------------------------

\SetTitle{\huge{Computational challenges in genome wide association studies: data processing, variant annotation and epistasis}}
\SetAuthor{Pablo Cingolani}
\SetDegreeType{PhD.}
\SetDepartment{School of Computer Science}
\SetUniversity{McGill University}
\SetUniversityAddr{Montreal, Quebec, Canada}
\SetThesisDate{July 2015}
\SetRequirements{A thesis submitted to McGill University in partial fulfillment of the requirements of the degree of Doctor of Philosophy}
\SetCopyright{Pablo Cingolani 2015}

%-----------------------------------------------------------------------------
% Document stats here
%-----------------------------------------------------------------------------

\begin{document}

% Create title page
\maketitle

%-----------------------------------------------------------------------------
% Input any special commands below
%-----------------------------------------------------------------------------

% Conditional expression for faster build during "development cycle"
\newif\ifthesis
\thesistrue       % Final thesis version: Uncomment this line and comment the next one
%\thesisfalse    % Development cycle: Uncomment this line and comment the previous one (faster LaTeX compile)

% Final version? 
% This is used to add items only for final submission
\newif\iffinal
%\finaltrue
\finalfalse

\ifthesis
	% Only show these sections when we are building the 'thesis' version

	\begin{romanPagenumber}{2}
	
	%-----------------------------------------------------------------------------
	% Acknowledgements:
	%   Among other acknowledgements, the student is required to declare the extent to which assistance (paid or unpaid) has 
	%   been given by members of staff, fellow students, research assistants, technicians, or others in the collection of materials 
	%   and data, the design and construction of apparatus, the performance of experiments, the analysis of data, and the 
	%   preparation of the thesis (including editorial help).
    %   In addition, it is appropriate to recognize the supervision and advice given by the thesis supervisor(s) and advisors.
	%-----------------------------------------------------------------------------		
	\SetAcknowledgeText{
I am most grateful to Mathieu Blanchette and Rob Sladek for the supervision of this thesis, their advice and guidance not only in professional issues, but also in all other fundamental aspects.
Many thanks to my PhD Committee: Jerome Waldispuhl, Doina Precup, Guillaume Bourque, and Derek Ruths for their helpful comments and suggestions. 
\\
I like to tank Douglas Ruden, Adrian Platts and Louis Letourneau for their insight and contributions to SnpEff and SnpSift projects.
\\
I am grateful to Mark McCarthy, John Blangero and Mike Boehnke, and David Altshuler for their leadership in the T2D consortia.
Special thanks to Pierre Fontanillas, Tanya Teslovich, Alisa Manning, Goo Jun, Anubha Mahajan, Jason Flannick, Andrew Morris, and Manuel Rivas for their helpful discussions that were instrumental in different aspects of this collaborative project.
\\
I thank Fiona Cunningham, Will McLaren, and Kai Wang for their contributions to the VCF variant annotation standard as well as Sarah Hunt for her efforts on the GA4GH annotations specification.
}
	\Acknowledge	

	%-----------------------------------------------------------------------------
	% Preface and contributions
	% In the case of collaborative work presented in either a standard format or manuscript-based thesis, there must be an 
	% explicit statement of the contributions of all parties, including the student, in the Preface of the thesis. 
   %  The Preface of a Doctoral thesis must also include a statement clearly indicating those elements of the thesis that 
   %  are considered original scholarship and distinct contributions to knowledge. 
	%-----------------------------------------------------------------------------		
	\SetPrefaceText{
	...the contributions of all parties, including the student....
	}
	\Preface	


	%-----------------------------------------------------------------------------
	%         English Abstract
	%-----------------------------------------------------------------------------
	
	\SetAbstractEnText{ 
Abundant genome sequence information from large cohorts of individuals can now be routinely obtained and this information is poised to ease the identification of genetic variations linked to complex disease. 
In this work, I investigate the computational and statistical challenges involved in the analysis of large genomic datasets and I tackle three different aspects of the analysis, each of them having very different characteristics.
First, in order to analyse large amounts of data from genomic studies we design a programming language, BigDataScript, that simplifies the creation of robust and scalable data analysis pipelines.
Second, we create genomic variant annotation and prioritization methods (SnpEff and SnpSift) that help to calculate putative genetic effects and estimate the genetic impact of variants.
Finally, we address the problem of finding associations between interacting genetic loci and disease by proposing a methodology that combines population-level genetic information with evolutionary information in order to increase the statistical power in epistatic genome wide association studies 
	}
	\AbstractEn
	
	%-----------------------------------------------------------------------------
	%         French Abstract
	%-----------------------------------------------------------------------------
	
	\SetAbstractFrText{ 
Il est aujourd'hui possible d'obtenir la s\'equence du g\'enome de grandes cohortes d'individus, et cette information est permet de faciliter l'identification de variations g\'en\'etiques li\'ees \`a des maladies complexes. 
Dans ma th\`ese, j'\'etudie les d\'efis informatiques et statistiques li\'es \`a l'analyse de grands ensembles de donn\'ees g\'enomiques. 
J'aborde trois aspects de l'analyse. 
Premi\`erement, afin d'analyser de grandes quantit\'es de donn\'ees provenant d'\'etudes g\'enomiques nous concevons un langage de programmation, BigDataScript, qui simplifie la cr\'eation de pipelines d'analyse de donn\'ees robustes et \'evolutives. 
Deuxi\`emement, nous cr\'eons deux m\'ethodes d'annotation et de classification de variantes g\'enomiques (SnpEff et SnpSift) qui aident \`a pr\'edire leur l'effet possible.
Enfin, nous abordons le probl\`eme de l'identification de liens entre les maladies g\'en\'etiques et les variantes qui les causent en proposant une m\'ethodologie qui combine l'information g\'en\'etique au niveau d'une la population avec informations \'evolutive afin d'augmenter la puissance statistique des \'etudes d'association consid\'erant les interactions \'epistatiques.
	}
	\AbstractFr
	
	%-----------------------------------------------------------------------------
	% Tables
	%-----------------------------------------------------------------------------
	
	%\renewcommand{\contentsname}{Table of contents}
	\tableofcontents 

	\LOFHeading{List of Figures and Tables}
	%\listoftables 
	\listoffigures
	
	\end{romanPagenumber}

\else
	% Skip all the previous sections
\fi

%-----------------------------------------------------------------------------
% Chapters
%-----------------------------------------------------------------------------

% McGill requires double spacing
\doublespacing

\ifthesis
	% Show all chapters on 'thesis' mode
	\include{chapters/intro}
	\include{chapters/intro_variants}
	\include{chapters/intro_epistasis}
	\include{chapters/intro_coevolution}
	\include{chapters/intro_roadmap}
	\include{chapters/bds}
	\include{chapters/snpeff}
	\include{chapters/gwas}
	\include{chapters/gwas_supp}
	\include{chapters/conclusions}
\else
	% Only show the chapter we are currently working on (faster LaTex build)
	\include{chapters/conclusions}
\fi

%-----------------------------------------------------------------------------
% Appendices
%-----------------------------------------------------------------------------

\ifthesis
	% Show all appendices (if any)
\else
	% Skip all appendices and bibliography
\fi
	
%-----------------------------------------------------------------------------
% Bibliography. You need to run the following 
% command: 
%       bibtex thesis_mcgill
%-----------------------------------------------------------------------------
\bibHeading{References}
\bibliographystyle{plain}
\bibliography{mcgilletd}		% File: mcgilletd.bib

\end{document}
